\begin{abstract}
When correlated measurements are naturally paired, through temporal repetition, matched subjects, or head-to-head competition, the variance of their difference depends on both the correlation structure and the relative magnitudes of the individual variances. Fisher's classical paired efficiency formula, $1/(1-\rho)$, captures this dependence only when variances are equal. We generalise this result to unequal variances through the Paired Efficiency Factor (PEF), $\eta = (1+\kappa)/(1+\kappa-2\sqrt{\kappa}\,\rho)$, where $\kappa$ denotes the variance ratio and $\rho$ the correlation coefficient. This formula is distribution-free and reduces to the classical case when $\kappa=1$.

Under a bivariate normal model, we derive a functional relationship between $\eta$ and the mutual information between the relative feature and a binary outcome. This relationship clarifies an apparent tension in machine learning practice: relative features can improve predictive performance even when $\eta<1$ (i.e., when relativisation increases variance), because information content and statistical efficiency measure distinct quantities. The two diverge most in competitive settings characterised by negative correlation and high variance asymmetry.

The primary empirical analysis uses key performance indicators (KPIs) from professional rugby union (United Rugby Championship ({URC}), $n=\PEFrugbyNgames$ games) and football (English Championship, $n=\PEFfootNgames$ games), where head-to-head competition provides natural pairing with both positive and negative correlation across KPIs. Supporting validation across four additional domains---pairwise oscillometric blood pressure readings from the {NHANES} examination component \parencite{nhanes2017pbxo}, paired transcript profiles from {GEO} series {GSE47462} \parencite{geoGSE47462,brunner2014earlybreast}, finance, and real industrial process telemetry ({OpenML} dataset 752; \parencite{openml752bosch})---shows domain-specific heterogeneity but overall alignment with the $(\kappa,\rho)$ taxonomy. The framework provides a four-quadrant guide for feature engineering decisions, supported by quadrant exemplars that demonstrate directional alignment between the $(\kappa,\rho)$ taxonomy and observed machine learning improvement when signal strength is controlled.



Comparing two entities through the difference of their measurements is a basic operation across the quantitative sciences, from paired clinical trials and financial benchmarking to head-to-head competition. When the paired measurements are correlated, pairing is the classical route to sharper inference, and Fisher's paired-efficiency result, $1/(1-\rho)$, quantifies the gain in terms of the correlation $\rho$. However, this result assumes the two measurements share a common variance---an assumption routinely violated in competitive and asymmetric settings, where practitioners also encounter a paradox: relative (difference) features often improve prediction even when pairing demonstrably increases variance, with no unified account of when relativisation helps or harms. Here we show that a single distribution-free quantity, the Paired Efficiency Factor (PEF) $\eta = (1+\kappa)/(1+\kappa-2\sqrt{\kappa}\,\rho)$---which generalises Fisher's formula to unequal variances through the variance ratio $\kappa$ and reduces to the classical case when $\kappa=1$---together with its link to predictive information, resolves this tension and organises feature-engineering decisions.

Under a bivariate normal model we derive a closed-form relationship between $\eta$ and the mutual information between the relative feature and a binary outcome, establishing that statistical efficiency and information content are distinct quantities: relative features can raise predictive power even when $\eta<1$ (i.e.\ when relativisation increases variance), and the two diverge most under negative correlation with high variance asymmetry. The $(\kappa,\rho)$ plane partitions into four quadrants, each with distinct feature-engineering guidance. Using key performance indicators (KPIs) from professional rugby union (United Rugby Championship ({URC}), $n=\PEFrugbyNgames$ games) and football (English Championship, $n=\PEFfootNgames$ games), where head-to-head competition provides natural pairing across both positive and negative correlation, quadrant exemplars confirm that the taxonomy predicts the direction of machine-learning gain once signal strength is controlled. Validation across four further domains---pairwise oscillometric blood pressure readings from the {NHANES} examination component \parencite{nhanes2017pbxo}, paired transcript profiles from {GEO} series {GSE47462} \parencite{geoGSE47462,brunner2014earlybreast}, finance, and real industrial process telemetry ({OpenML} dataset 752; \parencite{openml752bosch})---shows domain-specific heterogeneity but overall alignment with the $(\kappa,\rho)$ taxonomy. These results recast established practices---market-adjusted returns, paired clinical designs, manufacturing control charts, and relative sports metrics---as instances of one correlation-based principle, and provide a practical, distribution-free diagnostic for deciding when to relativise features in competitive prediction.

\end{abstract}
