\section{Results}
\label{sec:results}

\subsection{Theoretical Foundation: PEF--Information Content}

The theoretical relationship between $\eta$ and information content (\cref{eq:mi_closed}) was examined in four named scenarios from the theory-aligned idealised probit simulation (\cref{fig:si_idealised_stratified}; \texttt{run\_pef\_idealised\_probit\_sim.m}) under assumptions (A1)--(A2), with $n=5{,}000$ per cell, 50 Monte Carlo trials, and five-fold logistic cross-validation matching the empirical protocol. The simulation's data-generating process, estimands, and parameter-recovery checks are specified in full in Supplementary~Note~S2.

\begin{table}[t]
  \centering
  \caption{Scenario-level confirmation from the idealised probit simulation (A1)--(A2).}
  \label{tab:scenarios}
  \begin{tabular}{@{}lccccc@{}}
    \toprule
    \textbf{Scenario} & $\kappa$ & $\rho$ & $\eta$ & $I(X;Y)$ (bits) & ML impr.$^{\dagger}$ \\
    \midrule
    High competitive dynamics      & 2.0  & $-0.3$  & 0.780 & 0.029 & 9.4\% \\
    Moderate competitive dynamics  & 1.2  & $-0.15$ & 0.870 & 0.044 & 8.2\% \\
    Environmental dynamics         & 1.1  & 0.4     & 1.665 & 0.087 & 8.6\% \\
    Balanced (Fisher baseline)     & 1.0  & 0.0     & 1.000 & 0.056 & 7.9\% \\
    \bottomrule
  \end{tabular}

  \medskip
  \begin{minipage}{0.95\linewidth}\small
    $^{\dagger}$Mean five-fold cross-validated accuracy improvement (relative minus absolute features) across 50 Monte Carlo trials per scenario cell ($\delta/\sigma_{\mathrm{A}}=1$). Standard errors are $\approx$0.8--1.0 percentage points (see \texttt{idealised\_probit\_scenarios.csv}).
  \end{minipage}
\end{table}

On the admissible $(\kappa,\rho)$ grid at fixed $\delta/\sigma_{\mathrm{A}}=1$, analytic $\eta$ and $I(X;Y)$ from \cref{eq:mi_closed} are strongly correlated ($r\approx 0.87$ across 16 grid cells in the idealised simulation). The four scenarios in \cref{tab:scenarios} span the quadrant space under controlled (A1)--(A2) conditions; the first two illustrate the efficiency--power tension ($\eta<1$ with positive mean ML improvement), consistent with \cref{sec:eff_power_theory}.

\subsection{Primary Analysis: Sports KPIs}

The sports KPIs span all four quadrants (\cref{fig:pef_landscape}). Rugby yielded mean $\eta=\PEFrugbyMeanEta$ (95\% CI: $[\PEFrugbyCIlo,\PEFrugbyCIhi]$, \PEFrugbySuccess\% success rate across \PEFrugbyNkpis{} KPI studies). Football yielded mean $\eta=\PEFfootMeanEta$ (95\% CI: $[\PEFfootCIlo,\PEFfootCIhi]$, \PEFfootSuccess\% success across \PEFfootNkpis{} KPI studies). Representative exemplars illustrate the quadrant distribution in \cref{fig:pef_landscape}; per-KPI season-to-season stability of $(\kappa,\rho)$ positions appears in Supplementary~\cref{fig:si_kpi_labelled}.

KPIs with positive correlation (e.g., kicks from hand, carries) fell into Quadrants~1--2 and yielded $\eta>1$. KPIs with negative or near-zero correlation (e.g., turnovers won, penalties conceded) fell into Quadrants~3--4 and yielded $\eta\le 1$, yet several still showed positive ML improvement, consistent with the efficiency--power tension described in \cref{sec:eff_power_theory}.

\subsection{Quadrant-Specific Performance}

Empirical performance by quadrant:

\begin{table}[t]
  \centering
  \caption{Empirical performance by quadrant.}
  \label{tab:quad_perf}
  \begin{tabular}{@{}lcccc@{}}
    \toprule
    \textbf{Quadrant} & \textbf{Mean $\eta$} & \textbf{95\% CI} & \textbf{Success$^{\dagger}$} & \textbf{Mean ML impr.} \\
    \midrule
    Q1 (high $\kappa$, positive $\rho$) & \PEFqOneMeanEta   & $[\PEFqOneCIlo,\PEFqOneCIhi]$   & 100\% & \PEFqOneMLimpr\% \\
    Q2 (low $\kappa$, positive $\rho$)  & \PEFqTwoMeanEta   & ---                             & 95\%  & \PEFqTwoMLimpr\% \\
    Q3 (low $\kappa$, negative $\rho$)  & \PEFqThreeMeanEta & ---                             & 45\%  & \PEFqThreeMLimpr\% \\
    Q4 (high $\kappa$, negative $\rho$) & \PEFqFourMeanEta  & $[\PEFqFourCIlo,\PEFqFourCIhi]$ & 78\%$^{\ddagger}$  & \PEFqFourMLimpr\% \\
    \bottomrule
  \end{tabular}

  \medskip
  \begin{minipage}{0.95\linewidth}\small
    $^{\dagger}$Success in Q1--Q3: proportion of KPI studies with $\eta>1$ and $p<0.05$ (two-sided). $^{\ddagger}$Success in Q4 is redefined as observed ML improvement~$>0$, since $\eta>1$ is structurally impossible in this regime ($\kappa>1$, $\rho<0$); the 78\% figure reflects the proportion of Q4 KPI studies for which relative features yielded a positive accuracy gain.
  \end{minipage}
\end{table}

In Quadrant~4, mean $\eta=\PEFqFourMeanEta$ (indicating increased variance from pairing) yet mean ML improvement was \PEFqFourMLimpr\%, consistent with the analysis in \cref{sec:eff_power_theory}.

\subsection{Efficiency--Power Alignment Across Quadrants}
\label{sec:eff_power}

The four quadrants differ not only in mean $\eta$ but in the relationship between efficiency and predictive gain. Three supplementary analyses characterise this structure.

\textbf{Idealised simulation under (A1)--(A2).} On the theory-aligned probit grid (Supplementary~\cref{fig:si_idealised_stratified}; $n=5{,}000$, 50 trials, five-fold CV), all 16 out of 16 Quadrant~4 cells with $\eta<1$ produced positive mean ML improvement. At fixed $\delta/\sigma_{\mathrm{A}}$, $\eta$ and $I(X;Y)$ are strongly correlated (within-slice $r\approx0.87$ at the empirical median $\delta/\sigma_{\mathrm{A}}=\PEFmedianDeltaRatio$). However, the pooled-across-$\delta/\sigma_{\mathrm{A}}$ correlation is negative ($r=-0.70$), because signal strength shifts the $I(X;Y)$ level independently of the $(\kappa,\rho)$ position. The practical consequence is that iso-$\eta$ and iso-$I(X;Y)$ contours diverge once $\delta/\sigma_{\mathrm{A}}$ varies across KPIs (Supplementary~\cref{fig:si_iso_eta_I}).

\textbf{Quadrant distribution.} Of the \PEFtotalStudies{} sports KPI studies, 77 (68\%) lie in Q3 --- the negative-correlation regime where competitive dynamics dominate and absolute features are typically preferred. This reflects a structural feature of professional invasion-game sport: high-volume counts (passes, pressures, duels) are anti-correlated across opponents. Because the dataset is Q3-heavy, any aggregate statistic pooled across all KPIs primarily characterises this dominant regime rather than the mechanism across the full quadrant space. The exemplar approach (\cref{tab:exemplars}) directly addresses this by selecting one KPI per quadrant, allowing the taxonomy's directional predictions to be assessed without the distributional confound.

\textbf{Quadrant~4 and the Bayes bound.} Bootstrap confidence intervals on exemplar KPIs confirm that $\eta<1$ is estimated with low uncertainty for the Quadrant~4 sports KPIs studied (Supplementary~\cref{fig:si_bootstrap_exemplars}). Despite this, several Quadrant~4 KPIs yielded five-fold CV accuracy for relative features that exceeded the Gaussian Bayes-optimal bound on absolute features alone---a pattern consistent with the information content not being fully captured by the absolute-feature classifier (Supplementary~\cref{fig:si_q4_bayes_gap}). Together, these results support the interpretation in \cref{sec:eff_power_theory}: statistical efficiency and information content are related but distinct quantities, and the practical value of relativisation in competitive settings cannot be determined from $\eta$ alone.

\subsection{Quadrant Exemplars and Signal-Strength Context}
\label{sec:exemplars}

\Cref{fig:pef_ml} shows the empirical $\eta$--$\Delta\mathrm{ML}$ scatter for all \PEFtotalStudies{} sports KPI studies, with four annotated exemplars---one per quadrant---selected to illustrate how the $(\kappa,\rho)$ taxonomy predicts the direction of ML gain when signal strength is held approximately constant (\cref{tab:exemplars}).

\begin{table}[t]
  \centering
  \caption{Quadrant exemplars: four KPIs spanning the $(\kappa,\rho)$ parameter space.}
  \label{tab:exemplars}
  \begin{tabular}{@{}llccccc@{}}
    \toprule
    \textbf{Q} & \textbf{KPI (sport)} & $\kappa$ & $\rho$ & $\eta$ & $\delta/\sigma_{\mathrm{A}}$ & $\Delta\mathrm{ML}$ (\%) \\
    \midrule
    Q1 & Kick metres (rugby)              & 1.06 & $+0.65$ & 2.84 & 0.19 & $+4.8$ \\
    Q2 & Long balls (football)            & 0.91 & $+0.23$ & 1.30 & 0.16 & $+4.4$ \\
    Q3 & Passes (football)                & 0.84 & $-0.65$ & 0.61 & 0.32 & $-0.5$ \\
    Q4 & Goalkeeper long balls (football) & 1.00 & $-0.24$ & 0.81 & 0.26 & $+2.2$ \\
    \bottomrule
  \end{tabular}

  \medskip
  \begin{minipage}{0.95\linewidth}\small
    Two-season pooled estimates ($\kappa$, $\rho$, $\eta$, $\delta/\sigma_{\mathrm{A}}$) and five-fold cross-validated logistic improvement ($\Delta\mathrm{ML}$, relative minus absolute features). Signal strengths ($\delta/\sigma_{\mathrm{A}}\approx0.16$--$0.32$) are broadly comparable across exemplars. Q1 and Q2 show positive ML gain consistent with $\eta>1$; Q3 shows negative gain consistent with $\eta<1$; Q4 shows positive ML gain despite $\eta<1$, illustrating the efficiency--power tension (\cref{sec:eff_power_theory}).
  \end{minipage}
\end{table}

The Q4 exemplar (goalkeeper long balls, $\eta=0.81$) shows that relativisation can deliver a net ML benefit even when it statistically increases variance; here $\kappa\approx1.00$, so $\eta<1$ is driven by the negative correlation ($\rho=-0.24$) rather than by variance asymmetry, placing this KPI on the boundary between Quadrants~3 and~4. The Q3 exemplar (passes, football; $\rho=-0.65$, $\eta=0.61$) shows the complementary case: strong negative correlation places this KPI deep in the low-efficiency regime, and $\Delta\mathrm{ML}=-0.5\%$ confirms that absolute features are preferred, consistent with the taxonomy. Rucks won (rugby; Q2: $\kappa=0.82$, $\rho=+0.82$, $\eta=5.38$, $\delta/\sigma_{\mathrm{A}}=0.08$) provides a useful boundary illustration: despite the highest $\eta$ in the dataset, $\Delta\mathrm{ML}=0\%$ because the two teams are almost equally matched on average (negligible $\delta/\sigma_{\mathrm{A}}$). This confirms that even very high pairing efficiency confers no ML gain without sufficient signal strength.

\subsection{Supporting Validation: Cross-Domain Results}

Having established the mechanism in sports, where the single-KPI exemplars confirm the taxonomy's directional predictions, we now assess whether the same $(\kappa,\rho)$ structure recurs beyond sport. \Cref{tab:validation} summarises results across all six domains. The supporting domains illustrate how empirical $(\kappa,\rho)$ structures propagate into $\eta$ and success rates (here, proportion of units with $\eta>1$ and $p<0.05$ where hypothesis tests apply):
\begin{itemize}
  \item \textbf{Healthcare} \parencite{nhanes2017pbxo}: pooled $\eta=\PEFhealthMeanEta$ (bootstrap 95\% CI: $[2.448,2.622]$) across $N=\PEFhealthN$ {NHANES} participants pairing systolic and diastolic blood pressure from the same oscillometric measurement (Quadrant~2). Pulse pressure ($\mathrm{SBP}-\mathrm{DBP}$) is the clinically established relative feature.
  \item \textbf{Clinical genomics} \parencite{brunner2014earlybreast,geoGSE47462}: mean $\eta=\PEFgenomicsMeanEta$ ({SD} $=\PEFgenomicsSdEta$) across $\PEFgenomicsN$ genes ranked by paired-differential signal in {GSE47462}; $\PEFgenomicsSuccess\%$ of genes yielded $\eta>1$.
  \item \textbf{Finance}: mean $\eta=\PEFfinanceMeanEta$ ({SD} $=\PEFfinanceSdEta$) across $\PEFfinanceN$ {S\&P}~100 stocks relativised to daily {S\&P}~500 returns (2020--2023 pinned snapshot); $\PEFfinanceSuccess\%$ success rate.
  \item \textbf{Manufacturing} \parencite{openml752bosch}: mean $\eta=\PEFmfgMeanEta$ ({SD} $=\PEFmfgSdEta$) on {Bosch} {OpenML} 752 {CNC} cycles ($N=\PEFmfgN$ axis-wise pairings). Approximately $\PEFmfgSuccess\%$ of pairings yielded $\eta>1$.
\end{itemize}
Summary statistics for all six domains are collected in \cref{tab:validation}; the sports landscape with supporting-domain overlay markers appears in \cref{fig:pef_landscape}. Detailed per-study breakdowns for the non-sports domains are not separately visualised given the cross-sectional nature of those datasets.
